---
title: "Read counts to DGE, Part I"
author: "Friederike Dündar, ABC"
output:
  pdf_document: default
  html_document: default
  toc: true
editor_options: 
  chunk_output_type: console
---

This script will show you how to:
 
* Read in `featureCounts` results into R.
* Use `DESeq2` not (!) for DGE (yet), but to:
  - normalize read counts for differences in sequencing depth
  - transform reads to the log2 scale including variance reduction
* Accompany each step by exploratory plots.

You can generate an html document out of this entire script by clicking the `Knit HTML` button in RStudio.

```{r settingUp, warning=FALSE, echo=TRUE, message=FALSE}
options(stringsAsFactors = FALSE) # this will change a global setting, but just for this session

library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE,cache=FALSE) # tuning knitr output
```

## featureCounts

We aligned five samples for the WT and SNF2 condition, respectively.
You can find those files here: `~/mat/precomputed/results_alignment`.

>How can you check which command was used to generate those `BAM` files?

```{r featureCounts, eval=FALSE, echo = TRUE, engine='bash'}
## on the command line (!)
mkdir class/read_counts
cd class/read_counts
REF_DIR=~/class2019/mat/referenceGenomes/S_cerevisiae/

 # reads for yeast samples counted on the meta-feature level
~/class2019/mat/software/subread-1.6.0-Linux-x86_64/bin/featureCounts \
	-a ${REF_DIR}/Saccharomyces_cerevisiae.R64-1-1.81.gtf \
	-o featCounts_genes.txt \
	 ~/mat/precomputed/results_alignment/*bam 
```

Let's read the result file into R, i.e. download the table from [our website](http://chagall.med.cornell.edu/RNASEQcourse/featureCounts_result.txt) or
use the `scp` command to download the table that you generated on the server to
your local machine.

Loading additional libraries:

```{r}
library(ggplot2) # for making plots
library(magrittr) # for "pipe"-like coding in R
```

Reading the featureCounts result table into R:

```{r reading_in}
# reading in featureCounts output
readcounts <- read.table("~/Downloads/featCounts_genes.txt",
                          header=TRUE)
head(readcounts)
```

We will use the `DESeq2` package to normalize the samples for differences in their sequencing depths.
To make full use of `DESeq2`'s functions, we will have to turn the `data.frame` that we just generated via `read.table` into a specific R object that was defined by
the authors of `DESeq2`.

```{r install_bioconductor_pckgs, eval=FALSE, message = FALSE}
# not available via install.packages(), but through bioconductor
# installation should not be done everytime the Rmd file is compiled,
# hence the eval=FALSE code chunk setting to turn off its evaluation
BiocManager::install("DESeq2")
```

```{r warning=FALSE, message=FALSE}
library(DESeq2)
```

We will have to generate a `DESeqDataSet`; what is needed for this can be found 
out via `?DESeqDataSetFromMatrix`.
The help indicates that we need two tables: `countData` and `colData`.

* `colData`:  `data.frame` with all the variables you know about your samples, e.g., experimental condition, the type, and date of sequencing and so on. Its `row.names` should correspond to the unique sample names.
* `countData`: should contain a matrix of the actual values associated with the genes and samples. Is equivalent to `assay()`. Conveniently, this is almost exactly the format of the `featureCounts` output.

__Preparing the count matrix for the DESeq2DataSet class:__

```{r countMatrix_prep}
# give meaningful and legible sample names
## the original names of the sample columns are something like:
## 'X.home.classadmin.mat.precomputed.results_alignment.WT_5_Aligned.sortedByCoord.out.bam'
orig_names <- names(readcounts)
## here, we use two regular expressions with the gsub function
## 1. ".*alignment\\." --> any character (.*) occurring before an instance of
## 'alignment' followed by a dot (alignment\\.) should be replaced with nothing ("")
## 2. then, we also replace every instance of '_Aligned' and all subsequent 
## characters (.*) with nothing ("")
names(readcounts) <- gsub(".*alignment\\.", "" ,orig_names) %>% gsub("_Aligned.*", "", .)
```

```{r}
# gene IDs should be stored as row.names
row.names(readcounts) <- gsub("-", ".", readcounts$Geneid)
```
```{r}
# exclude the columns without read counts (columns 1 to 6 contain additional
# info such as genomic coordinates) 
readcounts <- readcounts[,-c(1:6)]
```

Always check your data set after you manipulated it!

```{r}
str(readcounts)
```

```{r}
head(readcounts)
```

In addition to the read counts, we need to add some more information about the samples, e.g. which condition and replicate they represent.
According to `?colData`, that type of "metadata" should be supplied in a `data.frame`,
where the _rows_ directly match the _columns_ of the count data.

Here's how this could be generated in `R` matching the `readcounts` `data.frame` we already have:

```{r making_colData}
sample_info <- DataFrame(condition = gsub("_[0-9]+", "", names(readcounts)),
                          row.names = names(readcounts) )
sample_info
```

```{r}
str(sample_info)
```

Let's generate the `DESeqDataSet`:

```{r DESeqDataSet, warning=FALSE}
DESeq.ds <- DESeqDataSetFromMatrix(countData = readcounts,
                              colData = sample_info,
                              design = ~ condition)
DESeq.ds
head(counts(DESeq.ds))
```

How many reads were counted for each sample ( = library sizes)?

```{r eval=TRUE, echo=TRUE}
colSums(counts(DESeq.ds))
```
```{r eval=TRUE, echo=TRUE}
colSums(counts(DESeq.ds)) %>% barplot
```

Remove genes with no reads.

```{r eval = TRUE}
keep_genes <- rowSums(counts(DESeq.ds)) > 0
dim(DESeq.ds)
```
```{r}
DESeq.ds <- DESeq.ds[ keep_genes, ]
dim(DESeq.ds)
```

```{r}
counts(DESeq.ds) %>% str
assay(DESeq.ds) %>% str
```

Now that we have the data in the right format, we can start using `DESeq2`'s functions, e.g. `estimateSizeFactors()` for calculating a factor that will be 
used to correct for sequencing depth differences.

```{r sizeFactors}
DESeq.ds <- estimateSizeFactors(DESeq.ds)
sizeFactors(DESeq.ds)
```

Check section 5.1.2 of the course notes to see the code for calculating the size
factors yourself with base R functions.

```{r sizeFactor_vs_librarySizes, eval=TRUE, echo=TRUE}
plot(sizeFactors(DESeq.ds), colSums(counts(DESeq.ds)))
```

The read counts normalized for sequencing depth can be accessed via
`counts(DESeq.ds, normalized = TRUE)`.

Let's check whether the normalization helped to adjust global differences between the samples.

```{r boxplots_untransformed, fig.width = 10, fig.height = 5}
# setting up the plotting layout
par(mfrow=c(1,2))
counts.sf_normalized <- counts(DESeq.ds, normalized=TRUE)

# adding the boxplots
boxplot(counts.sf_normalized, main = "SF normalized")
boxplot(counts(DESeq.ds), main = "read counts only")
```

We can't really see anything. 
It is usually helpful to *transform* the normalized read counts to bring them onto more similar scales.

>To see the influence of the sequencing depth normalization, make two box plots
of log2(read counts) - one for unnormalized counts, the other one for normalized
counts (exclude genes with zero reads in all samples).

```{r boxplots_logReadCounts, fig.width = 10, fig.height = 5, warning=FALSE}
par(mfrow=c(1,2)) # to plot the two box plots next to each other
boxplot(log2(counts(DESeq.ds)), notch=TRUE,
        main = "Non-normalized read counts\n(log-transformed)",
        ylab="read counts")
boxplot(log2(counts(DESeq.ds, normalize= TRUE)), notch=TRUE,
        main = "Size-factor-normalized read counts\n(log-transformed)",
        ylab="read counts") 
```


### Understanding more properties of read count data

Characteristics we've touched upon so far:

* zeros can mean two things: no expression or no detection
* fairly large dynamic range

>Make a scatterplot of log normalized counts against each other to see how well 
the actual values correlate which each other per sample and gene.

```{r logReadCountTables}
# non-normalized read counts plus pseudocount
log.counts <- log2(counts(DESeq.ds, normalized = FALSE) + 1)
# instead of creating a new object, we could assign the values to a distinct matrix
# normalized and log2-transformed read counts
assay(DESeq.ds, "log.norm.counts") <- log2(counts(DESeq.ds, normalized=TRUE) + 1)
```

```{r scatterplots_logNormReadCounts, fig.width = 10, fig.height = 15}
par(mfrow=c(2,1)) 
DESeq.ds[, c("WT_1","WT_2")] %>% assay(.,  "log.norm.counts") %>% 
    plot(., cex=.1, main = "WT_1 vs. WT_2")
DESeq.ds[, c("SNF2_1","SNF2_2")] %>% assay(.,  "log.norm.counts") %>% 
    plot(., cex=.1, main = "SNF2_1 vs SNF2_2")
```

Every dot = one gene.

The fanning out of the points in the lower left corner (points below $2^5 = 32$) 
indicates that read counts correlate less well between replicates when they are low.

This observation indicates that the standard deviation of the expression levels 
may depend on the mean: the lower the mean read counts per gene, the higher the
standard deviation.

This can be assessed visually; the package `vsn` offers a simple function for this.

```{r vsn}
par(mfrow=c(1,1))
# generate the base meanSdPlot using sequencing depth normalized log2(read counts)
msd_plot <- vsn::meanSdPlot(assay(DESeq.ds,  "log.norm.counts"), 
                       ranks=FALSE, # show the data on the original scale
                       plot = FALSE) # return a ggplot2 object without printing it

# add a title and y-axis label to the ggplot2 object
msd_plot$gg + 
  ggtitle("Sequencing depth normalized log2(read counts)") +
  ylab("standard deviation") 
```

From the help for`meanSdPlot`: *The red dots depict the running median estimator* 
*(window-width 10 percent). If there is no variance-mean dependence, then the line*
*formed by the red dots should be approximately horizontal.*

The plot here shows that there is some variance-mean dependence for genes with 
low read counts.
This means that the data shows signs of _heteroskedasticity_.

Many tools expect data to be _homoskedastic_, i.e., all variables should have similar variances.

`DESeq2` offers two ways to shrink the log-transformed counts for genes with very 
low counts: `rlog` and `varianceStabilizingTransformation` (`vst`).

We'll use `rlog` here as it is an optimized method for RNA-seq read counts:
it transforms the size-factor-normalized read counts to the log2 scale while simultaneously minimizing 
the difference between samples for rows with small counts and taking differences 
between library sizes of the samples into account.
`vst` tends to depend a bit more on the size factors, but generally, both methods
should return similar results.

```{r VarianceStabilization, fig.width = 15, fig.height = 8}
DESeq.rlog <- rlog(DESeq.ds, blind = TRUE) # this actually generates a
# different type of object
# set blind = FALSE if the conditions 
# are expected to introduce strong differences in a large proportion of the genes
```


```{r rlog_vs_log2, fig.width = 15, fig.height = 8}
par(mfrow=c(1,2)) 
plot(assay(DESeq.ds[,1:2],  "log.norm.counts"), cex=.1,
     main = "size factor and log2-transformed")

# the rlog-transformed counts are stored in the accessor "assay"
plot(assay(DESeq.rlog)[,1],
     assay(DESeq.rlog)[,2],
     cex=.1, main = "rlog transformed",
     xlab = colnames(assay(DESeq.rlog[,1])),
     ylab = colnames(assay(DESeq.rlog[,2])) )
rlog.norm.counts <- assay(DESeq.rlog)
```

As you can see in the left plot the variance - that is higher for small read 
counts - is tightened significantly using `rlog`.
What does the mean-sd-plot show?

```{r vsn_meanSdPlots_rlog}
# rlog-transformed read counts
msd_plot <- vsn::meanSdPlot( rlog.norm.counts, ranks=FALSE, plot = FALSE)
msd_plot$gg + ggtitle("rlog transformation")
```

```{r, echo=TRUE, eval = FALSE}
# I don't want to save every time I compile; therefore eval=FALSE
save.image(file = "~/Documents/Teaching/2019_RNA-seq/Rclass.RData")
```

### Similarity assessments and clustering

`pcaExplorer` lets you interact with the DESeq2-based plots and analyses.
It has included hierarchical clustering of samples and PCA.

#### `pcaExplorer`

```{r eval=FALSE}
pcaExplorer::pcaExplorer(dds = DESeq.ds, dst = DESeq.rlog)
```


#### Sample clustering using Pearson correlation

The ENCODE consortium recommends that _"for messenger RNA, (...) biological replicates [should] display greater than 0.9 correlation for transcripts/features"_.

The Pearson correlation coefficient is a measure of the strength of the linear relationship between two variables and is often used to assess the similarity of RNA-seq samples in a pair-wise fashion.
It is defined as the **covariance of two variables divided by the product of their standard deviation**.

Mimicking `pcaExplorer`'s heatmap:

```{r }
corr_coeff <- cor(rlog.norm.counts, method = "pearson")
as.dist(1-corr_coeff, upper = TRUE) %>% as.matrix %>%
    pheatmap::pheatmap(., main = "Pearson correlation")
```

Just plot the dendrogram, comparing the effects of the `rlog` transformation.

```{r fig.width=10, fig.height=5}
par(mfrow=c(1,2))
# Pearson corr. for rlog.norm values
as.dist(1 - corr_coeff) %>% hclust %>% 
    plot( ., 
        labels = colnames(rlog.norm.counts),
        main = "rlog transformed read counts")

# Pearson corr. for log.norm.values
as.dist( 1 - cor(assay(DESeq.ds, "log.norm.counts"), 
    method = "pearson")) %>% hclust %>%
    plot( .,
        labels = colnames(assay(DESeq.ds, "log.norm.counts")), 
        main = "no rlog")
```


#### How to do the PCA yourself (see the "protocol" part of pcaExplorer!)

```{r pca, eval=FALSE}
rv <- rowVars(assay(DESeq.rlog)) # equivalent to rowVars(rlog.norm.counts)
top_variable <- order(rv, decreasing = TRUE)[seq_len(500)]
pca <- prcomp(t(assay(DESeq.rlog)[top_variable, ]))
head(pca$x)
```


